Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-05-14 paper/main.tex
attempt to add figs (which themselves still need a touchup)

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}
\usepackage{url}

\begin{document}
\firstpage{1}

@@

\abstract{\textbf{Motivation:} Current state-of-the-art tools for the \textit{de novo} annotation of 
genes in eukaryotic genomes have to be specifically fitted for each species and still often 
produce annotations that can be improved much further. The fundamental algorithmic 
architecture for these tools has remained largely unchanged for about two decades, limiting
learning capabilities. Here, we set out to improve the cross-species annotation of
genes from DNA sequence alone with the help of deep learning. The goal is to eliminate 
the dependency on a closely related gene model while also improving the predictive
quality in general with a fundamentally new architecture.\\
\textbf{Results:} We present Helixer, a framework for the development and usage of 
a cross-species deep learning model that improves significantly on performance and
generalizability when compared to more traditional methods. We evaluate our 
approach by building a single model for the annotation of 186 animal genomes as 
well as one for 51 plants. Our predictions are shown to be much less sensitive to
the length of the genome than those of a current state-of-the-art tool. We also 
present two novel post-processing techniques that each worked to further strengthen 
our annotations and show in-depth results of an RNA-Seq based comparison of our 
predictions.\\
\textbf{Availability:} The source code of this work is available at 
\url{https://github.com/weberlab-hhu/HelixerPrep} [change] under the GNU General 
Public License v3.0.\\
\textbf{Contact:} \href{andreas.weber@hhu.de}{andreas.weber@hhu.de}\\

\textbf{Supplementary information:} Supplementary data are available at 
\textit{Bioinformatics} online.}

@@

As an independent evaluation and to get a better qualitative understanding of
performance we compared Helixer predictions, Augustus predictions and the
reference to RNAseq coverage data. RNAseq was processed for three each of plants
and animals chosen because they had varied performance quality relative to Augustus. 
RNAseq coverage provides support for an 
exonic annotation (UTR or CDS), spliced coverages provides support for an 
annotation of intron, and neither coverage nor spliced coverage is expected 
for intergenic annotations. Looking at selected subsequences \ref{fig:examples} % and supplemental figures.
we identified cases where RNAseq supported both the annotations of the reference 
and Helixer (a), neither (b), the reference, but not Helixer (c),
and Helixer, but not the reference (d).

@@

Coverage and Spliced coverage were broken down by the confusion matrix (of both
Helixer vs Augustus and Helixer vs the reference) for all genomes \ref{}. % H vs A main, H vs R supp
Where tools were in agreement (far left), coverage and spliced coverage closely matched
expectations. Specifically for CDS:CDS and UTR:UTR most base pairs had some, and many
had moderate or high coverage. The same pattern was seen for intron:intron and spliced
coverage. Finally, intergenic:intergenic showed only a small fraction of base pairs with 
any of either coverage or spliced coverage. In all conflicts the amount of RNAseq support
fell between the cases where tools agreed, indicating that all options were at least
capable of finding weak-spots in the other annotations.
In cases of conflict, Helixer's predictions were substantially more consistent with
the RNAseq data than those of Augustus. Specifically, for base pairs where one tool said CDS and
the other intergenic or intron \ref{}, there was more coverage when 
Helixer said CDS. Similarly, for base pairs where one tool said intron and the other 
CDS or intergenic, there was more spliced coverage when Helixer said intron \ref{}. 
These patterns were consistent in direction but varied in magnitude in the individual species,
with the expception of {\it P. marinus} where Helixer and Augustus performed comparably
at differentiating CDS and introns.

@@

Finally, developent of a post-processing method
to go from base pair wise predictions to integrated predictions for whole 
transcripts at each loci could both further improve performance and would
greatly increase real world applications.

@@

HHU HPC-team, deNBI, Ebenhöh lab for computational (particularly GPU) resources.

>>>>

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}
\usepackage{url}
\usepackage{placeins}
\begin{document}
\firstpage{1}

@@

\abstract{
\textbf{Motivation:} Current state-of-the-art tools for the \textit{de novo} annotation of 
genes in eukaryotic genomes have to be specifically fitted for each species and still often 
produce annotations that can be improved much further. The fundamental algorithmic 
architecture for these tools has remained largely unchanged for about two decades, limiting
learning capabilities. Here, we set out to improve the cross-species annotation of
genes from DNA sequence alone with the help of deep learning. The goal is to eliminate 
the dependency on a closely related gene model while also improving the predictive
quality in general with a fundamentally new architecture.\\
\textbf{Results:} We present Helixer, a framework for the development and usage of 
a cross-species deep learning model that improves significantly on performance and
generalizability when compared to more traditional methods. We evaluate our 
approach by building a single model for the annotation of 186 animal genomes as 
well as one for 51 plants. Our predictions are shown to be much less sensitive to
the length of the genome than those of a current state-of-the-art tool. We also 
present two novel post-processing techniques that each worked to further strengthen 
our annotations and show in-depth results of an RNA-Seq based comparison of our 
predictions.\\
\textbf{Availability:} The source code of this work is available at 
\url{https://github.com/weberlab-hhu/HelixerPrep} [change] under the GNU General 
Public License v3.0.\\
\textbf{Contact:} \href{andreas.weber@hhu.de}{andreas.weber@hhu.de}\\  
\textbf{Supplementary information:} Supplementary data are available at 
\textit{Bioinformatics} online.
}

@@

\begin{figure*}[!h]
\label{fig:cov_example_main}
\centerline{\includegraphics[width=\textwidth]{images/cov_examples/cov_example_main}}
\caption{high low accuracty by high low RNAseq support for ref}
\end{figure*}

As an independent evaluation and to get a better qualitative understanding of
performance we compared Helixer predictions, Augustus predictions and the
reference to RNAseq coverage data. RNAseq was processed for three each of plants
and animals chosen because they had varied performance quality relative to Augustus. 
RNAseq coverage provides support for an 
exonic annotation (UTR or CDS), spliced coverages provides support for an 
annotation of intron, and neither coverage nor spliced coverage is expected 
for intergenic annotations. Looking at selected subsequences (Figure \ref{fig:cov_example_main}) % and supplemental figures.
we identified cases where RNAseq supported both the annotations of the reference 
and Helixer (a), neither (b), the reference, but not Helixer (c),
and Helixer, but not the reference (d).

@@

\begin{figure*}[!h]
\label{fig:average_vs_augustus}
\includegraphics[width=5.6cm]{images/cov_examples/average_vs_augustus}  % should be 8.6 when floats are sorted out right
\caption{A caption}
\end{figure*}

Coverage and Spliced coverage were broken down by the confusion matrix (of both
Helixer vs Augustus and Helixer vs the reference) for all genomes (Figure \ref{fig:average_vs_augustus}). % H vs A main, H vs R supp
Where tools were in agreement (far left), coverage and spliced coverage closely matched
expectations. Specifically for CDS:CDS and UTR:UTR most base pairs had some, and many
had moderate or high coverage. The same pattern was seen for intron:intron and spliced
coverage. Finally, intergenic:intergenic showed only a small fraction of base pairs with 
any of either coverage or spliced coverage. In all conflicts the amount of RNAseq support
fell between the cases where tools agreed, indicating that all options were at least
capable of finding weak-spots in the other annotations.
In cases of conflict, Helixer's predictions were substantially more consistent with
the RNAseq data than those of Augustus. Specifically, for base pairs where one tool said CDS and
the other intergenic or intron \ref{}, there was more coverage when 
Helixer said CDS. Similarly, for base pairs where one tool said intron and the other 
CDS or intergenic, there was more spliced coverage when Helixer said intron \ref{}. 
These patterns were consistent in direction but varied in magnitude in the individual species,
with the expception of {\it P. marinus} where Helixer and Augustus performed comparably
at differentiating CDS and introns.

@@

Finally, developent of a post-processing method
to go from base pair wise predictions to integrated predictions for whole 
transcripts at each loci could both further improve performance and would
greatly increase real world applications.
\FloatBarrier

@@

HHU HPC-team, deNBI, Ebenh{\"o}h lab for computational (particularly GPU) resources.
